Efficient pedigree recording for fast population genetics simulation

### 2018-01-14 forwards_paper.tex
more explanation in complexity section

% I may need to cite my student's t-shirt for the following line.
Figures \ref{fig:runtimes_selection} and \ref{fig:relative_speedup_selection} show that this method
can dramatically improve simulation performance in practice --
but, how does it perform in theory?
% The simulation results shown in Figures \ref{fig:runtimes_selection} and \ref{fig:relative_speedup_selection} show that
% pedigree tracking greatly speeds up forward-time simulations.  Further, the run times with pedigree tracking become
% near-linear in $\rho$ (Figure \ref{fig:runtimes_selection}).
In this section, we explore how the required computational effort and storage space scale
in a Wright-Fisher model with recombination.

%% Memory reduction thanks to simplification
Consider a simulation of a Wright-Fisher population of $N$ haploid individuals
using Algorithm~\algref{W} for $T$ generations,
with simplify interval $s = 1$ so that redundant information is removed after every generation.
Since every chromosome inherits material from both parents,
without simplification this would produce tables of
$NT$ nodes and
$2NT$ edges.
Suppose that $T$ is large enough that all samples coalesce within the simulation with high probability
($T = 20N$, say, giving a total space requirement of $60N^2$).
After simplification, we are left with the tree sequence describing the history
of only the current generation of $N$ individuals.
If no coalescent events involve more than two lineages, this tree sequence has $2N-2$ edges to describe the leftmost tree;
and each time the marginal tree changes along the sequence,
four edges end and four new edges begin
(except for changes affecting the root, which require fewer; see \citet{kelleher2016efficient}).
Coalescent theory tells us that
the expected total length of the edges of a marginal tree is approximately $2N\log(N)$.
Since we have taken one crossover per individual per generation (as is typical for a single chromosome),
and we are measuring tree lengths in units of generations,
this is the mean number of ancestral recombination events that occur somewhere on the marginal trees.
Any such ancestral recombination might change the marginal gene tree;
if it does, it will add four new edges.
Not all such recombinations actually change the tree topology,
so at most we expect $8N\log(N)$ new edges after those required to describe the first tree.
Similarly, not every new edge derives from a never-before-seen node,
but the number of nodes is at most equal to the number of edges plus the sample size.
The space complexity after simplification is therefore $O(N + N\log N)$, since
we need $O(N)$ space to store the first marginal tree, and $O(1)$ space
for each of the $O(N \log N)$ trees that follow. Thus, we need $O(N^2)$ space
to store the complete history of the simulation, but only $O(N + N \log N)$
space to store the history that is relevant to the extant population.

%% Progressive memory reduction as a function of T
\citet{padhukasahasram2008exploring} obtained impressive speedups relative to contemporary simulators
by retaining only 8 generations of history.
This begs the question: how fast does the information in the pedigree become irrelevant?
Suppose instead that we have run a simulation of $2N$ individuals for $T$ generations,
begun with no prior history.
One way to answer this question is to ask:
How many of the resulting $4NT$ edges are required after simplification?
As above, the expected number of edges is bounded by $2N-2$ plus four times the mean marginal tree length.
Roughly, the length of time for which a marginal tree has $k$ tips is
$2N/(k(k-1)) = 2N(1/(k-1) - 1/k)$ generations,
and so the time to go from $N$ to $n$ tips is $2N(1/(n-1) - 1/(N-1))$.
This implies that after running for time $T$ we expect to
have around $n(T)$ roots, where $n(T) \approx 1 + 2N/(T+2)$.
The total tree length over this time is
$2 N \sum_{k=n(T)+1}^{N-1} 1/k$, which
% is approximately
% \begin{align*}
%     2 N \log\left( \frac{N}{1 + 2 N/(T+2)} \right) .
% \end{align*}
leads to an upper bound on the number of edges of
\begin{align}
    \label{eqn:edge_bound}
    2 N \left( 1 + 4 \log\left( \frac{NT}{T + 2 N} \right)\right) .
\end{align}
These calculations hold up quite well in practice,
as shown in Figure~\ref{fig:simplify_complexity}.
% by comparison to
% the number of edges actually required for 50 independent Wright--Fisher simulations,
% as a function of time (obtained from Algorithm~\algref{W} with $s=1$)
% is shown in Figure~\ref{fig:simplify_complexity}, compared to this prediction.

@@

Since each mutation is stored only as a single row in the mutation table,
and at most one row in the site table, the space required for $M$ mutations is $O(M)$;
combined with the $O(N \log N)$ storage for edges and nodes of a simplified tree sequence above,
this implies that the full results of a simulation of $N$ individuals
having $M$ mutations can be stored in $O(N \log N + M)$ space.
In simulations of whole chromosomes, the number of mutations per chromosome per generation is $O(1)$,
so $M$ will typically be $O(N \log N)$ as well.

@@

In this article, we applied our methods for storing trees to the problem of pedigree recording in a forward-time
simulation.  However, the method applies to any simulation scheme generating nodes and edges.  For example, one could
use the methods described here to generate succinct tree sequences under coalescent processes not currently implemented
in \msprime{}, such as the coalescent with gene conversion \citep{Wiuf2000-rc} or using the structured coalescent to
model various forms of natural selection \citep{Kaplan1988-in,Kaplan1989-rt,Braverman1995-gn}.  For such models, one
could in principle generate tables of nodes and edges to be simpified in \msprime{}.  The resulting succinct tree
sequence object would be in the same format as those generated by \msprime{}'s built-in simulate function, and therefore
compatible with exising methods for downstream analyses.

>>>>

% I may need to cite my student's t-shirt for the following line.
Figures \ref{fig:runtimes_selection} and \ref{fig:relative_speedup_selection} show that this method
can dramatically improve simulation performance in practice --
but, how does it perform in theory?
% The simulation results shown in Figures \ref{fig:runtimes_selection} and \ref{fig:relative_speedup_selection} show that
% pedigree tracking greatly speeds up forward-time simulations.  Further, the run times with pedigree tracking become
% near-linear in $\rho$ (Figure \ref{fig:runtimes_selection}).
In this section, we explore how the required computational effort and storage space scale
in a Wright-Fisher model with recombination.
Each edge in a tree sequence represents a segment of genome
that has been inherited across a certain number of generations, given by the length of the edge.
Each inheritance of each segment of genome across a generation
provides an opportunity for mutation and for recombination (which changes the tree).
For this reason, the key quantity we use is the total ``area'' of the tree sequence,
i.e., the product of segment length (left minus right coordinates) 
and edge length (difference in birth times between parent and child),
summed across all edges.
This area is also equal to the sum of the total lengths of all marginal trees
(i.e., the trees describing inheritance at each position on the genome),
so we can compute total area as sequence length
multiplied by the mean area of a given marginal tree,
which is well-described by coalescent theory.
Similar arguments to those below go back to \citet{watterson1975number},
who explicitly addresses smaller order corrections relevant to whole-population genealogies
of the Wright-Fisher model.
For more on coalescent theory, see \citet{wakeley2005coalescent}.

%% Memory reduction thanks to simplification
Consider a simulation of a Wright-Fisher population of $N$ haploid individuals
using Algorithm~\algref{W} for $T$ generations,
with simplify interval $s = 1$ so that redundant information is removed after every generation.
Since every chromosome inherits material from both parents,
without simplification this would produce tables of
$NT$ nodes and $2NT$ edges.
Suppose that $T$ is large enough that all samples coalesce within the simulation with high probability
($T = 20N$, say, giving a total space requirement of $60N^2$).
After simplification, we are left with the tree sequence describing the history
of only the current generation of $N$ individuals.
If no coalescent events involve more than two lineages, this tree sequence has $2N-2$ edges to describe the leftmost tree;
and each time the marginal tree changes along the sequence,
four edges end and four new edges begin
(except for changes affecting the root, which require fewer; see \citet{kelleher2016efficient}).
Coalescent theory tells us that
the expected total length of the edges of a marginal tree is approximately $2N\log(N)$.
Since we have taken one crossover per individual per generation (as is typical for a single chromosome),
and we are measuring tree lengths in units of generations,
this is the mean number of ancestral recombination events that occur somewhere on the marginal trees.
Any such ancestral recombination might change the marginal gene tree;
if it does, it will add four new edges.
Not all such recombinations actually change the tree topology,
so at most we expect $8N\log(N)$ new edges after those required to describe the first tree.
Similarly, not every new edge derives from a never-before-seen node,
but the number of nodes is at most equal to the number of edges plus the sample size.
The space complexity after simplification is therefore $O(N + N\log N)$, since
we need $O(N)$ space to store the first marginal tree, and a constant amount of space
for each of the $O(N \log N)$ trees that follow. Thus, we need $O(N^2)$ space
to store the complete history of the simulation, but only $O(N \log N)$ % $O(N + N \log N)$
space to store the history that is relevant to the extant population.

%% Progressive memory reduction as a function of T
\citet{padhukasahasram2008exploring} obtained impressive speedups relative to contemporary simulators
by retaining only 8 generations of history.
This begs the question: how fast does the information in the pedigree become irrelevant?
Suppose instead that we have run a simulation of $2N$ individuals for $T$ generations,
begun with no prior history.
One way to answer this question is to ask:
How many of the resulting $4NT$ edges are required after simplification?
As above, the expected number of edges is bounded by $2N-2$
plus four times the sum of the lengths of all marginal trees,
because new edges appear with most recombinations that occur on any marginal tree.
Therefore, we need to compute the mean total length of all branches in a coalescent tree
up until time $T$ (or the common ancestor, whichever comes first).
Since $k$ lineages coalesce at rate $\binom{k}{2}/N$,
the expected length of time for which a marginal tree has $k$ branches is
$2N/(k(k-1)) = 2N(1/(k-1) - 1/k)$ generations.
Since there are $k$ branches for this time,
this period is expected to contribute $2N/(k-1)$ generations to the total tree length.
By summing over $n < k \le N$,
the $N$ tips of a tree are expected to descend from only $n$ lineages
around $2N(1/(n-1) - 1/(N-1))$ generations ago.
Inverting this relationship between time and number of roots
implies that a marginal tree cut $T$ units of time ago
is expected to have around $n(T)$ roots, where $n(T) \approx 1 + 2N/(T+2)$.
The total tree length over this time is
$2 N \sum_{k=n(T)+1}^{N-1} 1/k$, which
% is approximately
% \begin{align*}
%     2 N \log\left( \frac{N}{1 + 2 N/(T+2)} \right) .
% \end{align*}
leads to an upper bound on the number of edges of
\begin{align}
    \label{eqn:edge_bound}
    2 N \left( 1 + 4 \log\left( \frac{NT}{T + 2 N} \right)\right) .
\end{align}
These calculations hold up quite well in practice,
as shown in Figure~\ref{fig:simplify_complexity}.
% by comparison to
% the number of edges actually required for 50 independent Wright--Fisher simulations,
% as a function of time (obtained from Algorithm~\algref{W} with $s=1$)
% is shown in Figure~\ref{fig:simplify_complexity}, compared to this prediction.

@@

Since each mutation is stored only as a single row in the mutation table,
and at most one row in the site table, the space required for $M$ mutations is $O(M)$;
combined with the $O(N \log N)$ storage for edges and nodes of a simplified tree sequence above,
this implies that the full results of a simulation of $N$ individuals
having $M$ mutations can be stored in $O(N \log N + M)$ space.
In simulations of whole chromosomes, there are typically a small, bounded number of mutations per chromosome per generation,
so $M$ will be $O(N \log N)$ as well.

@@

In this article, we applied our methods for storing trees to the problem of pedigree recording in a forward-time
simulation.  However, the method applies to any simulation scheme generating nodes and edges.  For example, one could
use the methods described here to generate succinct tree sequences under coalescent processes not currently implemented
in \msprime{}, such as the coalescent with gene conversion \citep{Wiuf2000-rc}, using the structured coalescent to
model various forms of natural selection \citep{Kaplan1988-in,Kaplan1989-rt,Braverman1995-gn},
or the coalescent within a known pedigree.  For such models, one
could in principle generate tables of nodes and edges to be simpified in \msprime{}.  The resulting succinct tree
sequence object would be in the same format as those generated by \msprime{}'s built-in simulate function, and therefore
compatible with exising methods for downstream analyses.
